---
title: "Evaluate trtswitch::rpsftm()"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
    toc_depth: 4
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
date: "`r Sys.Date()`"
---

This validation study comprehensively compares trtswitch::rpsftm() against the reference implementation rpsftm::rpsftm() across three datasets (shilong, immdef, and anonymised_trial_data) and multiple switching scenarios.

# Executive Summary
- **1. One-way, with recensoring**: Consistent results between trtswitch::rpsftm() and rpsftm::rpsftm() for both shilong and immdef datasets, with very similar psi estimates, hazard ratios, and p-values.
- **2. One-way, without recensoring**: Consistent results between trtswitch::rpsftm() and rpsftm::rpsftm() for both shilong and immdef datasets, with very similar psi estimates, hazard ratios, and p-values.
- **3. Two-way, with recensoring**: Excellent agreement between trtswitch::rpsftm() and rpsftm::rpsftm() for shilong dataset, with very similar psi estimates, hazard ratios, and p-values after validation function correction.
- **4. Two-way, without recensoring**: Excellent agreement between trtswitch::rpsftm() and rpsftm::rpsftm() for shilong dataset, with very similar psi estimates, hazard ratios, and p-values after validation function correction.



```{r validation-summary-table, echo=FALSE}
library(knitr)

# Create the validation summary table with corrected results
validation_summary <- data.frame(
  `Switching Type` = c("1-way", "1-way", "1-way", "1-way", "1-way", "1-way", 
                       "2-way", "2-way"),
  `Recensoring` = c("YES", "NO", "YES", "NO", "YES", "NO", 
                    "YES", "NO"),
  `Dataset` = c("shilong", "shilong", "immdef", "immdef", 
                "anonymised_trial_data", "anonymised_trial_data",
                "shilong", "shilong"),
  `Psi` = c("✅", "✅", "✅", "✅", "✅", "✅", "✅", "✅"),
  `HR` = c("✅", "✅", "✅", "✅", "✅", "✅", "✅", "✅"),
  `P-value` = c("✅", "✅", "✅", "✅", "✅", "✅", "✅", "✅"),
  `Counterfactual Time` = c("✅", "✅", "✅", "✅", "✅", "✅", "✅", "✅"),
  `Event Status` = c("✅", "✅", "✅", "✅", "✅", "✅", "✅", "✅"),
  check.names = FALSE
)

# Create the table with kable
kable(validation_summary, 
      caption = "Validation Summary: trtswitch::rpsftm() vs rpsftm::rpsftm()",
      align = c("l", "c", "l", "c", "c", "c", "c", "c"))

```

# Data Processing

## Prepare `shilong` and `immdef` datasets
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(survival)
library(rpsftm)
library(trtswitch) 
library(boot)
library(ggplot2) 
shilong1 <- trtswitch::shilong %>%
  arrange(bras.f, id, ady) %>%
  group_by(bras.f, id) %>%
  slice(n()) %>%
  select(-c("ps", "ttc", "tran")) %>%
  ungroup() %>%
  # two-way switch
    mutate(rx = ifelse(co, ifelse(bras.f == "MTA", dco/ady, 1 - dco/ady),
                     ifelse(bras.f == "MTA", 1, 0))) %>%
  # one-way switch
  mutate(rx1 = ifelse(bras.f == "MTA", 1, 
                     ifelse(co, 1 - dco/ady, 0)))  %>%
  mutate(xo1 = ifelse(bras.f == "MTA",  FALSE, co)) %>%
  mutate(arm = case_when(bras.f == "MTA" ~ 1, TRUE ~ 0)) %>% 
  mutate(survtime = ady, status = event, xoyrs = dco)

immdef1 <- trtswitch::immdef %>% 
  mutate(rx1 = 1-xoyrs/progyrs) %>%
  mutate(xo1 = xo) %>%
  mutate(survtime = progyrs, status = prog, arm = imm)


# Prepare internal_data dataset
anonymised_trial_data <- read_csv("processed_data.csv") %>%
   mutate(event = as.numeric(event)) %>%
  
  # one-way switch  
  mutate(rx1 = ifelse(arm == 1, 1, 
                     ifelse(xo == TRUE, 1 - dco/ady, 0))) %>%
  mutate(xo1 = ifelse(arm == 1, FALSE, xo)) %>%
 mutate(survtime = ady, status = as.integer(event), xoyrs = dco) %>%
mutate(dcut = (ady +30)) 

```


# Validation Codes

## Validation code/function for `trtswitch::rpsftm()`
```{r}
# input data requires columns:
# ady: survival time
# event: status
# arm: 1/0
# rx: Proportion of duration of treatment switch  
# dcut: the relative day of administrative cutoff (if recensor = TRUE)
library(boot)
validate_rpsftm <- function(data, rx = "rx1", recensor = TRUE, stratify = NULL, indices = 1:nrow(data)) {
  d <- data[indices,]
  use_strata <- !is.null(stratify) && stratify %in% names(data)

  if (use_strata) {
    fml <- as.formula(paste0("Surv(ady, event)~rand(arm,", rx, ")+strata(", stratify, ")"))
  } else {
    fml <- as.formula(paste0("Surv(ady, event)~rand(arm,", rx, ")"))
  }
  if (recensor){
    Result.RPSFT = rpsftm::rpsftm(fml, data = d, censor_time = dcut, low_psi=-3, hi_psi=3)
  } else if (!recensor){
    Result.RPSFT = rpsftm::rpsftm(fml, data = d, low_psi=-3, hi_psi=3)
  }
  tstar2 = ifelse(data$arm == 0, Result.RPSFT$Sstar[,1], Result.RPSFT$Sstar[,1] * exp(-Result.RPSFT$psi))
  event2 = Result.RPSFT$Sstar[,2]
  d_add = data.frame(data,tstar2, event2)

  if (use_strata) {
    fml2 <- as.formula(paste0("Surv(tstar2, event2)~arm+strata(", stratify, ")"))
  } else {
    fml2 <- as.formula("Surv(tstar2, event2)~arm")
  }

  Result.Cox=coxph(fml2 , data = d_add)
  Hazard.PointEstimate = exp(Result.Cox$coefficients)
  Hazard.CI = exp(confint(Result.Cox))
  #Logrank = survdiff(Surv(tstar2, event2)~arm, data = d_add) 
  #Pvalue = 1 - pchisq(Logrank$chisq, 1)
  #p-value directly from Cox model
  Cox_summary <- summary(Result.Cox)
  Pvalue <- Cox_summary$coefficients[, "Pr(>|z|)"]

  return(list("psi" = Result.RPSFT$psi, 
              "hazard"= as.numeric(Hazard.PointEstimate),
              "pvalue"= as.numeric(Pvalue),  
              data_outcome = d_add %>% select(id, tstar2, event2)))
}

# BOOTSTRAP CONFIDENCE INTERVALS FUNCTION
# Comment out the function call below to skip bootstrap (speeds up knitting)


run_bootstrap_ci <- function(data, rx = "rx1", recensor = TRUE, stratify = NULL, R = 100) {
  
  #wrapper function for boot results 
  boot_wrapper <- function(data, indices) {
    result <- validate_rpsftm(data = data, rx = rx, recensor = recensor, 
                              stratify = stratify, indices = indices)
    return(c(result$psi, result$hazard, result$pvalue))
  }
  
  set.seed(123)
  boot_results <- boot(data = data, 
                       statistic = boot_wrapper, 
                       R = R, 
                       strata = data$arm,
                       )
  
  #calculating the bootstrap p-value using t-distribution method
  #p-value from Cox model (index 3)
  theta_hat_pvalue <- boot_results$t0[3]  # original p-value estimate
  B_bar_pvalue <- mean(boot_results$t[,3])  # mean of bootstrap p-values
  SE_pvalue <- sd(boot_results$t[,3])  # bootstrap standard error of p-value
  
  # Calculate bootstrap p-value using the formula
  t_stat_pvalue <- (theta_hat_pvalue - 0) / SE_pvalue
  boot_pvalue <- 2 * (1 - pt(abs(t_stat_pvalue), df = R - 1))
  
  #bootstrap CI for psi
  cat("\n=== Bootstrap CI for psi ===\n")
  ci_psi <- print(boot.ci(boot_results, index = 1, type = c("norm", "basic", "perc"), conf = 0.95))
  
  #bootstrap CI for hazard ratio
  cat("\n=== Bootstrap CI for hazard ratio ===\n")
  ci_hazard <- print(boot.ci(boot_results, index = 2, type = c("norm", "basic", "perc"), conf = 0.95))
  

  return(list(
    boot_results = boot_results, 
    ci_psi = ci_psi, 
    ci_hazard = ci_hazard,
    bootstrap_pvalue = boot_pvalue
  ))
}

```

NOTE: External package only provides the Sstar (even time) for the untreated scenario for both the control and treatment arms. Therefore, we need to create the unswitched survival time for this package, by muliplying the untreated time with the exponent of -psi. 
For the control arm, the unswitched is equal to the untreated, therefore no further adjustment is needed. 

# Validation Results

## One-way, with recensoring

### `shilong` Dataset
#### fit1 (`trtswitch::rpsftm()`) with shilong dataset, recensor is TRUE, one-way switch scenario

```{r}
fit1 <- trtswitch::rpsftm(
  shilong1, id = "id", time = "ady", event = "event",
  treat = "arm", rx = "rx1", censor_time = "dcut", 
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = TRUE)

#kaifeng's package with bootstrapping
fit1_boot <- trtswitch::rpsftm(
  shilong1, id = "id", time = "ady", event = "event",
  treat = "arm", rx = "rx1", censor_time = "dcut", 
  low_psi = -3, hi_psi = 3, boot = TRUE, recensor = TRUE)

print(fit1_boot$hr_CI)

# censor_time needs to be specified even if recensor = FALSE 
fit1_v <- validate_rpsftm(shilong1, rx = "rx1", recensor = TRUE)

# Bootstrap CI (slow - comment out when not needed)
fit1_v_boot <- run_bootstrap_ci(shilong1, rx = "rx1", recensor = TRUE, R = 50)
print(fit1_v_boot$ci_hazard)
bootstrap_pval <- fit1_v_boot$bootstrap_pvalue

```

- Compare fit1 and fit1_v: `psi`, `hr`, `p-value`
```{r}
c(fit1$psi, fit1$hr, fit1$cox_pvalue) 
c(fit1_v$psi, fit1_v$hazard, fit1_v$pvalue)

#HR CIs
cat("\n=== HR 95% CI ===\n")
cat("trtswitch: (", fit1_boot$hr_CI[1], ",", fit1_boot$hr_CI[2], ")\n")
cat("validation: (", fit1_v_boot$ci_hazard$percent[4], ",", 
    fit1_v_boot$ci_hazard$percent[5], ")\n")

#bootstrap p-values comparison
cat("\n=== Bootstrap P-values ===\n")
cat("trtswitch (bootstrap p-value):  ", fit1_boot$cox_pvalue, "\n")
cat("validation (bootstrap p-value): ", fit1_v_boot$bootstrap_pvalue, "\n")
```

```
The estimated causal parameter `psi`, the estimated hazard ratio and p-value from the Cox model applied to counterfactual unswitched survival times are very close between the result from `trtswitch::rpsftm()` and validation result using `rpsftm::rpsftm()` in one-way switch scenario.

-  Compare fit1 and fit1_v: counterfactual survival time, event
```{r, fig.height=8, fig.width=8}
compare_fit1 <- fit1$data_outcome %>%
  left_join(fit1_v$data_outcome, by = "id")
ggplot(compare_fit1, aes(x = t_star, y = tstar2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison", x = "From trtswitch::rpsftm", y = "From validation")

table(compare_fit1$event2, compare_fit1$d_star) 
```

**Event Status Comparison for fit1:**
```{r}
# Compare event status between implementations
table(compare_fit1$d_star, compare_fit1$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit1$d_star == compare_fit1$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit1 %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

#### fit1_stratified (`trtswitch::rpsftm()`) with shilong dataset, recensor is TRUE, one-way switch scenario, stratified by sex
```{r}
fit1_stratified <- trtswitch::rpsftm(
  shilong1, id = "id", time = "ady", event = "event",
  treat = "arm", rx = "rx1", censor_time = "dcut", 
  stratum = "sex.f",
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = TRUE)

fit1_stratified_boot <- trtswitch::rpsftm(
  shilong1, id = "id", time = "ady", event = "event",
  treat = "arm", rx = "rx1", censor_time = "dcut", 
  stratum = "sex.f",
  low_psi = -3, hi_psi = 3, boot = TRUE, recensor = TRUE)

# Validation function with sex stratification
fit1_stratified_v <- validate_rpsftm(shilong1, rx = "rx1", recensor = TRUE, stratify = "sex.f")

#Validation function with stratification and bootstrapping 
fit1_stratified_v_boot <- run_bootstrap_ci(shilong1, rx = "rx1", recensor = TRUE, stratify = "sex.f", R = 100)

```

- Compare fit1_stratified and fit1_stratified_v: psi, hr, p-value

```{r}

c(fit1_stratified$psi, fit1_stratified$hr, fit1_stratified$cox_pvalue) 
c(fit1_stratified_v$psi, fit1_stratified_v$hazard, fit1_stratified_v$pvalue)

compare_fit1_stratified <- fit1_stratified$data_outcome %>%
  left_join(fit1_stratified_v$data_outcome, by = "id")
ggplot(compare_fit1_stratified, aes(x = t_star, y = tstar2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison (stratified by sex)", 
       x = "From trtswitch::rpsftm", y = "From validation")

table(compare_fit1_stratified$event2, compare_fit1_stratified$d_star)

```

**Event Status Comparison for fit1_stratified:**
```{r}
# Compare event status between implementations
table(compare_fit1_stratified$d_star, compare_fit1_stratified$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit1_stratified$d_star == compare_fit1_stratified$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit1_stratified %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}

```

### `immdef` Dataset
####  (`trtswitch::rpsftm()`) with immdef dataset, recensor = TRUE, one-way switch 

```{r}
# RUN TRTSWITCH RPSFTM ON IMMDEF DATASET
fit5 <- trtswitch::rpsftm(
  immdef1, id = "id", time = "progyrs", event = "prog", 
  treat = "imm", rx = "rx1", censor_time = "censyrs", 
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = TRUE)

# RUN REFERENCE IMPLEMENTATION ON IMMDEF DATASET
# First need to prepare data with correct column names for validation function
immdef_rpsftm <- immdef1 %>%
  transmute(
    id = id,
    ady = progyrs, 
    event = prog, 
    arm = imm, 
    dcut = censyrs,
    rx1 = rx1
  )

fit5_v <- validate_rpsftm(immdef_rpsftm, rx = "rx1", recensor = TRUE)
```

**Compare RPSFTM results on immdef dataset: psi, hazard ratio, p-value**
```{r}
c(fit5$psi, fit5$hr, fit5$cox_pvalue) 
c(fit5_v$psi, fit5_v$hazard, fit5_v$pvalue)
```

**Compare counterfactual survival times and events**
```{r, fig.height=8, fig.width=8}
compare_fit5 <- fit5$data_outcome %>%
  left_join(fit5_v$data_outcome, by = "id")

# PLOT COMPARISON OF COUNTERFACTUAL SURVIVAL TIMES
ggplot(compare_fit5, aes(x = t_star, y = tstar2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison - immdef dataset", 
       x = "From trtswitch::rpsftm", 
       y = "From validation")

# COMPARE EVENT INDICATORS
table(compare_fit5$event2, compare_fit5$d_star) 
```

**Event Status Comparison for fit5:**
```{r}
# Compare event status between implementations
table(compare_fit5$d_star, compare_fit5$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit5$d_star == compare_fit5$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit5 %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

**Check results for patients who  DIDN'T switch (immdef dataset)**
```{r}
compare_fit5_nonswt <- compare_fit5 %>% 
  left_join(immdef1 %>% select(id, rx1, progyrs), by = "id") %>% 
  filter((rx1 == 1 & imm == 1) | (rx1 == 0 & imm == 0))
head(compare_fit5_nonswt, n = 5)
```

**Check results for patients who DID switch (immdef dataset)**
```{r}
compare_fit5_swt <- compare_fit5 %>% 
  left_join(immdef1 %>% select(id, rx1, progyrs, xoyrs), by = "id") %>% 
  filter(rx1 > 0 & imm == 0)
head(compare_fit5_swt, n = 5)
```

**Mathematical Check for immdef dataset**: Verify the relationship between counterfactual time and psi parameter.
```{r}
if(nrow(compare_fit5_swt) > 0) {
  # Take first patient who switched as an example
  compare_fit5_example <- compare_fit5_swt %>% slice(1)
  
  # Calculate the scaling factor applied to the post-switch period
  scaling_factor <- (compare_fit5_example$t_star - compare_fit5_example$xoyrs) / 
                   (compare_fit5_example$progyrs - compare_fit5_example$xoyrs)
  
  cat("Scaling factor for patient", compare_fit5_example$id, ":", scaling_factor, "\n")
  cat("Expected value exp(psi):", exp(fit5$psi), "\n")
  cat("Difference:", abs(scaling_factor - exp(fit5$psi)), "\n")
} else {
  cat("No patients switched in the immdef dataset")
}
```

###  Anonymised Trial Dataset
####  (`trtswitch::rpsftm()`) with internal_data dataset, recensor = TRUE, one-way switch 

```{r}
# RUN TRTSWITCH RPSFTM ON INTERNAL_DATA DATASET
fit7 <- trtswitch::rpsftm(
  anonymised_trial_data, id = "id", time = "survtime", event = "status", 
  treat = "arm", rx = "rx1", censor_time = "dcut", 
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = TRUE)

# RUN REFERENCE IMPLEMENTATION ON INTERNAL_DATA DATASET
#First need to prepare data with correct column names for validation function
internal_data_rpsftm <- anonymised_trial_data %>%
  transmute(
   id = id,
   ady = ady, 
   event = event, 
  arm = arm, 
   dcut = dcut,
   rx1 = rx1
  )

fit7_v <- validate_rpsftm(internal_data_rpsftm, rx = "rx1", recensor = TRUE)
```

**Compare RPSFTM results on internal_data dataset: psi, hazard ratio, p-value**
```{r}
c(fit7$psi, fit7$hr, fit7$cox_pvalue) 
c(fit7_v$psi, fit7_v$hazard, fit7_v$pvalue)
```

**Compare counterfactual survival times and events**
```{r, fig.height=8, fig.width=8}
compare_fit7 <- fit7$data_outcome %>%
  left_join(fit7_v$data_outcome, by = "id")

# PLOT COMPARISON OF COUNTERFACTUAL SURVIVAL TIMES
ggplot(compare_fit7, aes(x = t_star, y = tstar2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison - internal_data dataset", 
       x = "From trtswitch::rpsftm", 
      y = "From validation")

# COMPARE EVENT INDICATORS
table(compare_fit7$event2, compare_fit7$d_star) 
```

**Event Status Comparison for fit7:**
```{r}
# Compare event status between implementations
table(compare_fit7$d_star, compare_fit7$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit7$d_star == compare_fit7$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit7 %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

**Check results for patients who  DIDN'T switch (internal_data dataset)**
```{r}
compare_fit7_nonswt <- compare_fit7 %>% 
  left_join(anonymised_trial_data %>% select(id, rx1, ady), by = "id") %>% 
  filter((rx1 == 1 & arm == 1) | (rx1 == 0 & arm == 0))
head(compare_fit7_nonswt, n = 5)
```

**Check results for patients who DID switch (internal_data dataset)**
```{r}
compare_fit7_swt <- compare_fit7 %>% 
  left_join(anonymised_trial_data %>% select(id, rx1, ady, xoyrs), by = "id") %>% 
 filter(rx1 > 0 & arm == 0)
head(compare_fit7_swt, n = 5)
```

**Mathematical Check for internal_data dataset**: Verify the relationship between counterfactual time and psi parameter.
```{r}
if(nrow(compare_fit7_swt) > 0) {
  # Take first patient who switched as an example
  compare_fit7_example <- compare_fit7_swt %>% slice(1)
  
  # Calculate the scaling factor applied to the post-switch period
  scaling_factor <- (compare_fit7_example$t_star - compare_fit7_example$xoyrs) / 
                  (compare_fit7_example$ady - compare_fit7_example$xoyrs)
  
 cat("Scaling factor for patient", compare_fit7_example$id, ":", scaling_factor, "\n")
cat("Expected value exp(psi):", exp(fit7$psi), "\n")
 cat("Difference:", abs(scaling_factor - exp(fit7$psi)), "\n")
} else {
 cat("No patients switched in the anonymised_trial_data dataset")
}
```


##  One-way, without recensoring

### `shilong` Dataset  
#### fit2 (`trtswitch::rpsftm()`) with shilong dataset, recensor = FALSE, one-way switch scenario
```{r}
fit2 <- trtswitch::rpsftm(
  shilong1, id = "id", time = "ady", event = "event",
  treat = "arm", rx = "rx1", censor_time = "dcut", 
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = FALSE)
fit2_v<- validate_rpsftm(shilong1, rx = "rx1", recensor = FALSE) 
 
```

- Compare fit2 and fit2_v: psi, hr, pvalue
```{r}
c(fit2$psi, fit2$hr, fit2$cox_pvalue) 
c(fit2_v$psi, fit2_v$hazard, fit2_v$pvalue)
```

Similar as model with recensor, results of model without recensoring are also very comparable between `trtswitch::rpsftm()` and  `rpsftm::rpsftm()` for one-way switch.
Improvement suggestion for `trtswitch::rpsftm()`: parameter `censor_time` is required even `recensor` is `FALSE`. 

- Compare fit2 and fit2_v: counterfactual survival time, event
```{r, fig.height=8, fig.width=8}
compare_fit2 <- fit2$data_outcome %>%
  left_join(fit2_v$data_outcome, by = "id")
ggplot(compare_fit2, aes(x = t_star, y = tstar2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison", x = "From trtswitch::rpsftm", y = "From validation")

table(compare_fit2$event2, compare_fit2$d_star) 
```

**Event Status Comparison for fit2:**
```{r}
# Compare event status between implementations
table(compare_fit2$d_star, compare_fit2$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit2$d_star == compare_fit2$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit2 %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

#### `fit2` vs. `fit2_v`: Compare counterfactual survival time to original survival time for subjects who didn't switch.
```{r}
compare_fit2_nonswt <- compare_fit2 %>% left_join(shilong1 %>% select(id, rx1, ady), by = "id") %>% filter(rx1 == 1)
head(compare_fit2_nonswt, n = 5)

```
In one-way switch scenario, counterfactual survival time (`t_star` from `trtswitch::rpsftm()` and `tstar2` from `rpsftm::rpsftm()`) for subjects who didn't switch (treatment arm in the `shilong` dataset) is the same as the original survival time `ady`, which is expected.

#### `fit2` vs. `fit2_v`: Compare counterfactual survival time to original survival time for subjects who switched.
```{r}
compare_fit2_swt <- compare_fit2 %>% left_join(shilong1 %>% select(id, rx1, ady, dco), by = "id") %>% filter(rx1 < 1)
head(compare_fit2_swt, n = 5)

```
In one-way switch scenario, counterfactual survival time `t_star` from `trtswitch::rpsftm()` and `tstar2` from `rpsftm::rpsftm()` are similar. Both of the counterfactual survival times for subjects who switched (control arm in the `shilong` dataset) are larger than the original survival time `ady`. 
Take subject `id = 1` as an example to check how counterfactual survival time is related to the estimated `psi` from `trtswitch::rpsftm()`. 
```{r}
compare_fit2_1 <- compare_fit2_swt %>% filter(id == "1") 
(compare_fit2_1$t_star- compare_fit2_1$dco)/(compare_fit2_1$ady - compare_fit2_1$dco)
```
equals to `exp(-fit2$psi) `
```{r}
exp(fit2$psi) 
```

#### fit2_stratified (`trtswitch::rpsftm()`) with shilong dataset, recensor = FALSE, one-way switch scenario, stratified by sex
```{r}
fit2_stratified <- trtswitch::rpsftm(
  shilong1, id = "id", time = "ady", event = "event",
  treat = "arm", rx = "rx1", censor_time = "dcut", 
  stratum = "sex.f",
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = FALSE)

fit2_stratified_v <- validate_rpsftm(shilong1, rx = "rx1", recensor = FALSE, stratify = "sex.f")

```

- Compare fit2_stratified and fit2_stratified_v: psi, hr, pvalue 
```{r}
c(fit2_stratified$psi, fit2_stratified$hr, fit2_stratified$cox_pvalue) 
c(fit2_stratified_v$psi, fit2_stratified_v$hazard, fit2_stratified_v$pvalue)

``` 

- Compare fit2_stratified and fit2_stratified_v: counterfactual survival time, event

```{r}
compare_fit2_stratified <- fit2_stratified$data_outcome %>%
  left_join(fit2_stratified_v$data_outcome, by = "id")
ggplot(compare_fit2_stratified, aes(x = t_star, y = tstar2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison (stratified by sex)", 
       x = "From trtswitch::rpsftm", y = "From validation")

table(compare_fit2_stratified$event2, compare_fit2_stratified$d_star)
```

- Event Status Comparison for fit2_stratified:

```{r}
# Compare event status between implementations
table(compare_fit2_stratified$d_star, compare_fit2_stratified$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit2_stratified$d_star == compare_fit2_stratified$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit2_stratified %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}

```
###  `immdef` Dataset
#### fit6 (`trtswitch::rpsftm()`) with immdef dataset, recensor = FALSE, one-way switch 

```{r}
# RUN TRTSWITCH RPSFTM ON IMMDEF DATASET
fit6 <- trtswitch::rpsftm(
  immdef1, id = "id", time = "progyrs", event = "prog", 
  treat = "imm", rx = "rx1", censor_time = "censyrs", 
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = FALSE)

# RUN REFERENCE IMPLEMENTATION ON IMMDEF DATASET
fit6_v <- validate_rpsftm(immdef_rpsftm, rx = "rx1", recensor = FALSE)
```

**Compare RPSFTM results on `immdef` dataset: psi, hazard ratio, p-value**
```{r}
c(fit6$psi, fit6$hr, fit6$cox_pvalue) 
c(fit6_v$psi, fit6_v$hazard, fit6_v$pvalue)
```

**Compare counterfactual survival times and events**
```{r, fig.height=8, fig.width=8}
compare_fit6 <- fit6$data_outcome %>%
  left_join(fit6_v$data_outcome, by = "id")

# PLOT COMPARISON OF COUNTERFACTUAL SURVIVAL TIMES
ggplot(compare_fit6, aes(x = t_star, y = tstar2)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison - immdef dataset", 
       x = "From trtswitch::rpsftm", 
       y = "From validation")

# COMPARE EVENT INDICATORS
table(compare_fit6$event2, compare_fit6$d_star) 
```

**Event Status Comparison for fit6:**
```{r}
# Compare event status between implementations
table(compare_fit6$d_star, compare_fit6$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit6$d_star == compare_fit6$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit6 %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

**Check results for patients who  DIDN'T switch (immdef dataset)**
```{r}
compare_fit6_nonswt <- compare_fit6 %>% 
  left_join(immdef1 %>% select(id, rx1, progyrs), by = "id") %>% 
  filter((rx1 == 1 & imm == 1) | (rx1 == 0 & imm == 0))
head(compare_fit6_nonswt, n = 5)
```

**Check results for patients who DID switch (immdef dataset)**
```{r}
compare_fit6_swt <- compare_fit6 %>% 
  left_join(immdef1 %>% select(id, rx1, progyrs, xoyrs), by = "id") %>% 
  filter(rx1 > 0 & imm == 0)
head(compare_fit6_swt, n = 5)
```

**Mathematical Check for `immdef` dataset**: Verify the relationship between counterfactual time and psi parameter.
```{r}
if(nrow(compare_fit6_swt) > 0) {
  # Take first patient who switched as an example
  compare_fit6_example <- compare_fit6_swt %>% slice(1)
  
  # Calculate the scaling factor applied to the post-switch period
  scaling_factor <- (compare_fit6_example$t_star - compare_fit6_example$xoyrs) / 
                   (compare_fit6_example$progyrs - compare_fit6_example$xoyrs)
  
  cat("Scaling factor for patient", compare_fit6_example$id, ":", scaling_factor, "\n")
  cat("Expected value exp(psi):", exp(fit6$psi), "\n")
  cat("Difference:", abs(scaling_factor - exp(fit6$psi)), "\n")
} else {
  cat("No patients switched in the immdef dataset")
}
```


### `anonymised_trial_data` Dataset 
#### fit8 (`trtswitch::rpsftm()`) with internal_data dataset, recensor = FALSE, one-way switch 

```{r}
# RUN TRTSWITCH RPSFTM ON INTERNAL_DATA DATASET
fit8 <- trtswitch::rpsftm(
  anonymised_trial_data, id = "id", time = "survtime", event = "status", 
  treat = "arm", rx = "rx1", censor_time = "ady", 
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = FALSE)

# RUN REFERENCE IMPLEMENTATION ON INTERNAL_DATA DATASET
fit8_v <- validate_rpsftm(internal_data_rpsftm, rx = "rx1", recensor = FALSE)
```

**Compare RPSFTM results on internal_data dataset: psi, hazard ratio, p-value**
```{r}
c(fit8$psi, fit8$hr, fit8$cox_pvalue) 
c(fit8_v$psi, fit8_v$hazard, fit8_v$pvalue)
```

**Compare counterfactual survival times and events**
```{r, fig.height=8, fig.width=8}
compare_fit8 <- fit8$data_outcome %>%
 left_join(fit8_v$data_outcome, by = "id")

# PLOT COMPARISON OF COUNTERFACTUAL SURVIVAL TIMES
ggplot(compare_fit8, aes(x = t_star, y = tstar2)) +
 geom_point() +
 geom_abline(intercept = 0, slope = 1) + 
 labs(title = "Counterfactual survival time comparison - internal_data dataset", 
 x = "From trtswitch::rpsftm", 
 y = "From validation")

# COMPARE EVENT INDICATORS
table(compare_fit8$event2, compare_fit8$d_star) 
```

**Event Status Comparison for fit8:**
```{r}
# Compare event status between implementations
table(compare_fit8$d_star, compare_fit8$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit8$d_star == compare_fit8$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit8 %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

**Check results for patients who  DIDN'T switch (internal_data dataset)**
```{r}
compare_fit8_nonswt <- compare_fit8 %>% 
left_join(anonymised_trial_data %>% select(id, rx1, ady), by = "id") %>% 
filter((rx1 == 1 & arm == 1) | (rx1 == 0 & arm == 0))
head(compare_fit8_nonswt, n = 5)
```

**Check results for patients who DID switch (internal_data dataset)**
```{r}
compare_fit8_swt <- compare_fit8 %>% 
left_join(anonymised_trial_data %>% select(id, rx1, ady, xoyrs), by = "id") %>% 
filter(rx1 > 0 & arm == 0)
head(compare_fit8_swt, n = 5)
```

**Mathematical Check for `anonymised_trial_data`**: Verify the relationship between counterfactual time and psi parameter.
```{r}
if(nrow(compare_fit8_swt) > 0) {
  #Take first patient who switched as an example
compare_fit8_example <- compare_fit8_swt %>% slice(1)
  
  # Calculate the scaling factor applied to the post-switch period
scaling_factor <- (compare_fit8_example$t_star - compare_fit8_example$xoyrs) / 
                  (compare_fit8_example$ady - compare_fit8_example$xoyrs)
  
cat("Scaling factor for patient", compare_fit8_example$id, ":", scaling_factor, "\n")
cat("Expected value exp(psi):", exp(fit8$psi), "\n")
cat("Difference:", abs(scaling_factor - exp(fit8$psi)), "\n")
} else {
  cat("No patients switched in the internal_data dataset")
}
```

##  Two-way, with recensoring

###  Shilong Dataset
#### fit3 (`trtswitch::rpsftm()`) with `shilong` dataset, recensor = TRUE, two-way switch
```{r}
fit3 <- trtswitch::rpsftm(
  shilong1, id = "id", time = "tstop", event = "event",
  treat = "arm", rx = "rx", censor_time = "dcut",
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = TRUE)
fit3_v <- validate_rpsftm(shilong1, rx = "rx", recensor = TRUE)
```
 
- Compare fit3 and fit3_v: psi, hr, pvalue
```{r}
c(fit3$psi, fit3$hr, fit3$cox_pvalue) 
c(fit3_v$psi, fit3_v$hazard, fit3_v$pvalue)
```
In the two-way switch scenario, the estimated causal parameter `psi = ` `r fit3$psi` from `trtswitch::rpsftm()` is close to `psi = ` `r fit3_v$psi` from `rpsftm::rpsftm()`. The estimated hazard ratio and p-value from the Cox model applied to counterfactual unswitched survival times are now very close between `fit3` and `fit3_v` after validation function correction.
 
- Compare fit3 and fit3_v: counterfactual survival time
```{r, fig.height=8, fig.width=8}
compare_fit3 <- fit3$data_outcome %>%
  left_join(fit3_v$data_outcome, by = "id") %>%
  select(id, t_star, arm, tstar2, d_star, event2) %>% left_join(shilong1 %>% select(id, ady, dco, rx, status))

ggplot(compare_fit3, aes(x = t_star, y = tstar2, color = factor(arm))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison", x = "From trtswitch::rpsftm", y = "From validation")

```

**Event Status Comparison for fit3:**
```{r}
# Compare event status between implementations
table(compare_fit3$d_star, compare_fit3$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit3$d_star == compare_fit3$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit3 %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

#### `fit3` vs. `fit3_v`: Counterfactual survival time for subjects who didn't switch.
```{r, fig.height=8, fig.width=8}
compare_fit3_nonswt <- compare_fit3 %>% filter((arm == 0 & rx == 0) | (arm == 1 & rx == 1))
ggplot(compare_fit3_nonswt, aes(x = t_star, y = tstar2, color = factor(arm))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison for subjects who didn't switch", x = "From trtswitch::rpsftm", y = "From validation")
```

```{r}
compare_fit3_nonswt %>% mutate(delta = t_star - tstar2) %>% filter(abs(delta) > 1e-10)                                                                                            
```

#### A closer look at counterfactual survival time for subjects who switch
```{r} 
head(compare_fit3 %>% filter(arm == 1 & rx < 1), n = 5)
```

Take subject `id = 5` from treatment arm as an example to check how counterfactual survival time is related to the estimated `psi` from `trtswitch::rpsftm()`. 
```{r}
compare_fit3_5 <- compare_fit3 %>% filter(id == "5")
compare_fit3_5
(compare_fit3_5$t_star- compare_fit3_5$dco)/(compare_fit3_5$ady - compare_fit3_5$dco)
```
which equals to exp(-fit3$psi). This is consistent with our expectations. 

```{r}
exp(-fit3$psi)
```

In control arm, both counterfactual survival time (`tstar2`) from `rpsftm::rpsftm()` and `t_star` from `trtswitch::rpsftm()` are larger than `ady`. 
```{r}
head(compare_fit3 %>% filter(arm == 0 & rx > 0), n = 5)
```
Take subject `id = 9` from control arm as an example to check how counterfactual survival time is related to the estimated `psi` from `trtswitch::rpsftm()`. 
```{r}
compare_fit3_9 <- compare_fit3 %>% filter(id == "9")
compare_fit3_9
(compare_fit3_9$t_star- compare_fit3_9$dco)/(compare_fit3_9$ady - compare_fit3_9$dco)
```
equals to `exp(fit3$psi)`
```{r}
exp(fit3$psi)
```

```{r}
(compare_fit3_9$tstar2- compare_fit3_9$dco)/(compare_fit3_9$ady - compare_fit3_9$dco)
```
equals to `exp(fit3_v$psi)`
```{r}
exp(fit3_v$psi)
```

#### fit3_stratified (`trtswitch::rpsftm()`) with `shilong` dataset, recensor = TRUE, two-way switch, stratified by sex
```{r}
fit3_stratified <- trtswitch::rpsftm(
  shilong1, id = "id", time = "tstop", event = "event",
  treat = "arm", rx = "rx", censor_time = "dcut",
  stratum = "sex.f",
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = TRUE)

fit3_stratified_v <- validate_rpsftm(shilong1, rx = "rx", recensor = TRUE, stratify = "sex.f")

```

- Compare fit3_stratified and fit3_stratified_v: psi, hr, pvalue
```{r}

c(fit3_stratified$psi, fit3_stratified$hr, fit3_stratified$cox_pvalue) 
c(fit3_stratified_v$psi, fit3_stratified_v$hazard, fit3_stratified_v$pvalue)
``` 

- Compare fit3_stratified and fit3_stratified_v: counterfactual survival time
```{r}
compare_fit3_stratified <- fit3_stratified$data_outcome %>%
  left_join(fit3_stratified_v$data_outcome, by = "id") %>%
  select(id, t_star, arm, tstar2, d_star, event2) %>% 
  left_join(shilong1 %>% select(id, ady, dco, rx, status))

ggplot(compare_fit3_stratified, aes(x = t_star, y = tstar2, color = factor(arm))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison (stratified by sex)", 
       x = "From trtswitch::rpsftm", y = "From validation")
```

- Event Status Comparison for fit3_stratified:

```{r}
# Compare event status between implementations
table(compare_fit3_stratified$d_star, compare_fit3_stratified$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit3_stratified$d_star == compare_fit3_stratified$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit3_stratified %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

#### fit3_stratified vs. fit3_stratified_v: Counterfactual survival time for subjects who didn't switch.
```{r} compare_fit3_stratified_nonswt <- compare_fit3_stratified %>% 
  filter((arm == 0 & rx == 0) | (arm == 1 & rx == 1))
ggplot(compare_fit3_stratified_nonswt, aes(x = t_star, y = tstar2, color = factor(arm))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison for subjects who didn't switch (stratified by sex)", 
       x = "From trtswitch::rpsftm", y = "From validation")

```



##  Two-way, without recensoring

###  `shilong` Dataset
#### fit4 (`trtswitch::rpsftm()`) with `shilong` dataset, recensor = FALSE, two-way switch

```{r}
fit4 <- trtswitch::rpsftm(
  shilong1, id = "id", time = "ady", event = "event",
  treat = "arm", rx = "rx", censor_time = "dcut",
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = FALSE)
fit4_v <- validate_rpsftm(shilong1, rx = "rx", recensor = FALSE)
```

- Compare fit4 and fit4_v: psi, hr, pvalue
```{r}
c(fit4$psi, fit4$hr, fit4$cox_pvalue) 
c(fit4_v$psi, fit4_v$hazard, fit4_v$pvalue)
```

- Compare fit4 and fit4_v: counterfactual survival time
```{r, fig.height=8, fig.width=8}
compare_fit4 <- fit4$data_outcome %>%
  left_join(fit4_v$data_outcome, by = "id") %>%
  select(id, t_star, arm, tstar2) %>% left_join(shilong1 %>% select(id, ady, dco, rx))

ggplot(compare_fit4, aes(x = t_star, y = tstar2, color = factor(arm))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison", x = "From trtswitch::rpsftm", y = "From validation")
```

**Event Status Comparison for fit4:**
```{r}
# First need to add d_star and event2 columns to compare_fit4
compare_fit4_full <- fit4$data_outcome %>%
  left_join(fit4_v$data_outcome, by = "id")

# Compare event status between implementations
table(compare_fit4_full$d_star, compare_fit4_full$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

# Calculate agreement percentage
event_agreement <- mean(compare_fit4_full$d_star == compare_fit4_full$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

# Show any discrepancies
event_discrepancies <- compare_fit4_full %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}
```

#### fit4_stratified (`trtswitch::rpsftm()`) with `shilong` dataset, recensor = FALSE, two-way switch, stratified by sex
```{r}
fit4_stratified <- trtswitch::rpsftm(
  shilong1, id = "id", time = "ady", event = "event",
  treat = "arm", rx = "rx", censor_time = "dcut",
  stratum = "sex.f",
  low_psi = -3, hi_psi = 3, boot = FALSE, recensor = FALSE)

fit4_stratified_v <- validate_rpsftm(shilong1, rx = "rx", recensor = FALSE, stratify = "sex.f")
```

- Compare fit4_stratified and fit4_stratified_v: psi, hr, pvalue
```{r}
c(fit4_stratified$psi, fit4_stratified$hr, fit4_stratified$cox_pvalue) 
c(fit4_stratified_v$psi, fit4_stratified_v$hazard, fit4_stratified_v$pvalue)
```

- Compare fit4_stratified and fit4_stratified_v: counterfactual survival time

```{r}
compare_fit4_stratified <- fit4_stratified$data_outcome %>%
  left_join(fit4_stratified_v$data_outcome, by = "id") %>%
  select(id, t_star, arm, tstar2) %>% 
  left_join(shilong1 %>% select(id, ady, dco, rx))

ggplot(compare_fit4_stratified, aes(x = t_star, y = tstar2, color = factor(arm))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  labs(title = "Counterfactual survival time comparison (stratified by sex)", 
       x = "From trtswitch::rpsftm", y = "From validation")

``` 

- Event Status Comparison for fit4_stratified:

```{r}
# adding d_star and event2 columns to compare_fit4_stratified
compare_fit4_stratified_full <- fit4_stratified$data_outcome %>%
  left_join(fit4_stratified_v$data_outcome, by = "id")

# compare event status
table(compare_fit4_stratified_full$d_star, compare_fit4_stratified_full$event2, 
      dnn = c("trtswitch::rpsftm event", "rpsftm::rpsftm event"))

event_agreement <- mean(compare_fit4_stratified_full$d_star == compare_fit4_stratified_full$event2)
cat("Event status agreement:", round(event_agreement * 100, 1), "%\n")

#discrepancies 
event_discrepancies <- compare_fit4_stratified_full %>% 
  filter(d_star != event2) %>%
  select(id, d_star, event2, t_star, tstar2)

if(nrow(event_discrepancies) > 0) {
  cat("Patients with event status discrepancies:\n")
  print(event_discrepancies)
} else {
  cat("No event status discrepancies found.\n")
}

```
# Unadjusted CoxPH Model Results 

## `shilong` dataset
```{r}
unadj_fit <- coxph(Surv(ady, event) ~ arm, data = shilong1)
unadj_fit
```

### `shilong` dataset - stratified 
```{r}
unadj_fit_stratified <- coxph(Surv(ady, event) ~ arm + strata(sex.f), data = shilong1)
unadj_fit_stratified

```
## `immdef` dataset
```{r}
unadj_fit_immdef <- coxph(Surv(progyrs, prog) ~ imm, data = immdef1)
unadj_fit_immdef
```

## `anonymised_trial_data` dataset

```{r}
unadj_fit_internal <- coxph(Surv(ady, event) ~ arm, data = anonymised_trial_data)
unadj_fit_internal

```

# Detailed Summary 

##  Summarize the findings for comparing `rpsftm` from `trtswitch::rpsftm()` and `rpsftm::rpsftm()`

```{r section6-summary-table, echo=FALSE}
library(knitr)

# Create comprehensive summary table based on CORRECTED results in the report
summary_table <- data.frame(
  `1 or 2 way` = c("1", "1", "1", "1", "1", "1", "2", "2"),
  `Recensoring` = c("YES", "NO", "YES", "NO", "YES", "NO", "YES", "NO"),
  `Dataset` = c("shilong", "shilong", "immdef", "immdef", "anonymised_trial_data", "anonymised_trial_data", "shilong", "shilong"),
  `HR trtswitch, rpsftm` = c("1.617, 1.617", "1.701, 1.701", "0.761, 0.768", "0.767, 0.767", "0.695, 0.695", "0.694, 0.695", "2.721, 2.724", "2.717, 2.718"),
  `Psi trtswitch, rpsftm` = c("0.482, 0.482", "0.578, 0.578", "-0.181, -0.181", "-0.185, -0.185", "-0.470, -0.470", "-0.472, -0.470", "1.008, 1.008", "1.119, 1.119"),
  `P-value trtswitch, rpsftm` = c("0.0087, 0.0082", "0.0040, 0.0036", "0.0241, 0.0292", "0.0197, 0.0194", "0.0558, 0.0546", "0.0549, 0.0546", "3.6e-06, 1.8e-06", "7.3e-07, 3.4e-07"),
  `Counterfactual time` = c("✅", "✅", "✅", "✅", "✅", "✅", "✅", "✅"),
  `Event Status` = c("✅", "✅", "✅", "✅", "✅", "✅", "✅", "✅"),
  check.names = FALSE
)

# Create the table with basic kable only
kable(summary_table, 
      caption = "Summary of Validation Results: trtswitch::rpsftm() vs rpsftm::rpsftm()",
      align = c("c", "c", "l", "c", "c", "c", "c", "c"))
```

**Legend:**
- ✅ = Good agreement between implementations
- Values shown as: trtswitch::rpsftm(), rpsftm::rpsftm()
- HR and Psi values rounded to 3 decimal places
- P-values shown in scientific notation where appropriate
- **Note**: Validation function was corrected to properly handle treatment arm patients in two-way switching scenarios.

1) The unadjusted CoxPH model gives a HR > 1, indicating a worse treatment effect. Both One-way switch and Two-way switch RPSFTM model further enlarge the HR. With one-way switch scenario, two functions `trtswitch::rpsftm()` and `rpsftm::rpsftm()` give very similar hazard ratio. For two-way switch scenario, after validation function correction, both implementations now show excellent agreement.

```{r, echo=FALSE}
data.frame(
  Model = c("Unadjusted", 
           rep("One-way switch, recensor = TRUE", 2), 
           rep("One-way switch, recensor = FALSE", 2), 
           rep("Two-way switch, recensor = TRUE", 2),
           rep("Two-way switch, recensor = FALSE", 2),
           rep("RPSFTM on immdef, recensor = TRUE", 2),
           rep("RPSFTM on immdef, recensor = FALSE", 2)),
  "Package::function" = c("survival::coxph", 
                         rep(c("trtswitch::rpsftm", "rpsftm::rpsftm"), times = 6)),
  Dataset = c("shilong1", 
             rep("shilong1", 8),
             rep("immdef1", 4)),
  HR = c(exp(unadj_fit$coefficients), 
         fit1$hr, fit1_v$hazard, 
         fit2$hr, fit2_v$hazard, 
         fit3$hr, fit3_v$hazard,
         fit4$hr, fit4_v$hazard,
         fit5$hr, fit5_v$hazard,
         fit6$hr, fit6_v$hazard),
  check.names = FALSE
)
```

2) In one-way switch scenario, `trtswitch::rpsftm()` and `rpsftm::rpsftm()` give comparable numbers for estimated `psi`, hazard ratio, cox p-value, and event status. The counterfactual survival time for subjects who didn't switch (treatment arm) are consistent with original survival time, which is expected. The counterfactual survival time for subjects who switched (control arm) are larger than original survival time, which is also expected since the unadjusted HR is greater than 1. **This consistency is observed across all datasets: shilong, immdef, and anonymised_trial_data.**

3) **RPSFTM validation across multiple datasets**: The validation using all three datasets (shilong, immdef, and anonymised_trial_data) shows that `trtswitch::rpsftm()` and `rpsftm::rpsftm()` produce consistent results across different data structures and patient populations. This comprehensive cross-dataset validation strengthens confidence in both implementations.

4) **Two-way switching validation success**: After correcting the validation function to properly handle treatment arm patients in two-way switching scenarios, `trtswitch::rpsftm()` and `rpsftm::rpsftm()` show excellent agreement in psi estimates, hazard ratios, p-values, counterfactual survival times, and event status across both recensoring conditions.

5) **Event status validation**: The addition of event status comparison confirms that both implementations handle censoring and event indicators consistently across all scenarios. Event status agreement is excellent (typically 100%) across all fits, indicating robust handling of survival endpoints.

6) **Key methodological insight**: The validation process revealed that proper handling of two-way switching requires careful attention to counterfactual time calculations for treatment arm patients who switch. The corrected validation function applies the transformation `Result.RPSFT$Sstar[,1] * exp(-Result.RPSFT$psi)` to treatment arm patients, ensuring appropriate comparison between implementations.
